---
title: "Class Reference"
author: "Aamaly Hossain"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Consider this a personal guide to the commands and functions you will learn. In general, when you come across an R command or function that you want to remember, put it in here along with a description of what it does and when you'd use it.

### How to set the working directory

The command to set a working directory is setwd(). For example, this sets the working directory to a directory called data_journalism_2024_fall inside a directory called jour472 located inside your home directory (the ~ is equivalent of /Users/[youruser] on a Mac).

```{r}
#setwd("~/jour472/data_journalism_2024_fall")
```

### How to install and load the tidyverse

```{r}
#install.packages("tidyverse")
library(tidyverse)
```


```{r}
number <- 2
#<- assigns a value (2) to a variable (number) so it can be reused later.



number * number
#* multiplies values; here it squares number to return 4.



#install.packages("tidyverse")
#install.packages() 
#installs an external library so you can use its functions.



library(tidyverse)
#library() loads a library (like tidyverse) so its functions are available.



#umd_courses <- read_rds("umd_courses.rds")
#read_rds() loads an .rds file into R as a dataframe (here named umd_courses).



#glimpse(umd_courses)
#glimpse() shows variable names, types, and sample values for a dataframe.



#head(umd_courses)
#head() displays the first six rows of a dataframe.



umd_courses |>
  group_by(term) |>
  summarise(count_classes = n())
#"|>" pipes data forward, group_by() groups rows by a variable, and summarise() creates summaries (here n() counts rows).



arrange(desc(count_classes))
#arrange() sorts rows; desc() makes the order descending.



summarise(total_seats = sum(seats))
#sum() adds up all values in a column (here the total seats).


summarise(mean_seats = mean(seats),
         median_seats = median(seats))
#mean() calculates the average; median() finds the middle value.



summarise(minimum_seats = min(seats),
          maximum_seats = max(seats))
#min() finds the smallest value; max() finds the largest value.
```



```{r}
knitr::opts_chunk$set(echo = TRUE)
#knitr::opts_chunk$set() sets options for RMarkdown code chunks (here, echo = TRUE means code is shown in the output).


options(scipen=999)
#options(scipen=999) turns off scientific notation so large numbers display normally.


library(tidyverse)
#Loads the tidyverse, a collection of R packages for data manipulation and visualization.


library(lubridate)
#Loads the lubridate package, which makes it easier to work with dates and times.


#umd_courses <- read_rds("data/umd_courses.rds")
#read_rds() reads an RDS file (binary R data format) into a dataframe.


summary(umd_courses)
#summary() gives basic stats for each column (min, mean, max for numbers; counts for categories).


colnames(umd_courses)
#colnames() lists the names of all columns in a dataframe.


journalism_courses <- umd_courses |> filter(department == "Journalism")
#filter() keeps only rows meeting a condition (here, department is "Journalism").


courses_2025 <- umd_courses |> filter(term > 202500)
#Filters to only rows where the term is greater than 202500.


selected_journalism_courses <- journalism_courses |> select(id, title)
#select() keeps only the chosen columns (id and title) while keeping all rows.


theatre_seats_15 <- umd_courses |> filter(department == "Theatre") |> filter(seats >= 15)
#Multiple filter() calls can be chained to narrow rows step by step.


and_theatre_seats_15 <- umd_courses |> filter(department == "Theatre" & seats >= 15)
#Using & (AND) in a single filter() combines multiple conditions at once.


umd_courses |> filter(str_detect(title, "Shakespeare"))
#str_detect() finds rows where a string column (title) contains a given word or pattern.


#maryland_expenses <- read_csv("data/maryland_expenses.csv")
#read_csv() reads CSV files into a dataframe (from the readr package).


maryland_expenses <- maryland_expenses |> mutate(expenditure_date=mdy(expenditure_date))
#mutate() creates or updates columns; here it converts a character date into a proper date using mdy().


maryland_expenses |>
  mutate(month = floor_date(expenditure_date, "month"))
#floor_date() rounds a date down to the nearest unit (here, the first day of the month).


general_22 <- read_csv('data/md_gov_county.csv')
#Reads in county-level election results from CSV.


general_22 <- general_22 |>
  mutate(
    total_votes = cox + moore + lashar + wallace + write_ins,
    pct_moore = moore/total_votes
  )
#Creates new columns: total_votes adds up all candidate votes; pct_moore calculates Moore’s share.


mutate(pct_moore = (moore/total_votes)*100)
#Converts Moore’s share from a decimal to a percentage.


arrange(desc(pct_moore))
#arrange() sorts rows by a column; desc() puts them in descending order.


arrange(pct_moore)
#Same as above, but sorts ascending.


mutate(payee_upper = str_to_upper(payee_name))
#str_to_upper() converts all text in a column to uppercase for consistency.

#Other similar functions:

#str_to_lower() → makes text lowercase.

#str_to_sentence() → makes only the first letter of the first word uppercase.


mutate(
  state = case_when(
      str_detect(address, " Maryland ") ~ "MD",
      str_detect(address, " California ") ~ "CA",
      .default = NA
    )
)
#case_when() creates new values conditionally: if an address contains a state name, it assigns the abbreviation; otherwise it assigns NA.

group_by(state) |> summarize(total = sum(amount))
#group_by() clusters rows by state; summarize() collapses each group into totals (here, total spending).
```



```{r}
payments <- read_csv("data/file.csv", guess_max=10)
#read_csv() reads a CSV file; the guess_max argument tells R how many rows to look at when guessing column types.

problems(payments)
#Shows parsing issues (e.g., if R misguessed a column’s type when reading in the file).

glimpse(payments)
#Gives a compact look at the dataframe: column names, data types, and sample values.

group_by(`Vendor Zip`) |>
  filter(str_length(`Vendor Zip`) < 5) |>
  summarise(count=n()) |>
  arrange(desc(count))
#group_by() groups data by a variable.
#filter() keeps only rows meeting a condition.
#str_length() counts how many characters are in a string.
#summarise(count=n()) counts the rows in each group.
#arrange(desc(count)) sorts results in descending order.

md_grants_loans <- read_csv("data/file.csv")
#Reads another CSV into a dataframe.

group_by(`Fiscal Year`, Category) |> summarize(count = n()) |> arrange(`Fiscal Year`)
#Groups rows by fiscal year and category, counts them, then sorts by fiscal year.

arrange(Amount)
#Sorts all rows by the Amount column (ascending by default).

install.packages("janitor")
#Installs the janitor package, used for cleaning messy data.

library(janitor)
#Loads the janitor package.

clean_names()
#From janitor: standardizes column names (all lowercase, underscores instead of spaces).

rename(source = grantor)
#Renames a column (grantor) to a new name (source); new name goes first.

mutate(source = str_to_upper(source), grantee = str_to_upper(grantee), description = str_to_upper(description))
#mutate() changes or creates columns; str_to_upper() makes all text uppercase.

get_dupes()
#From janitor: finds duplicate rows in the dataframe.

distinct()
#Removes duplicate rows, keeping only unique ones.

mutate(zip5 = str_sub(zip_code, start=1L, end=5L))
#Creates a new column (zip5) that takes only the first 5 characters from the zip_code column.

mutate(zip5 = case_when(
  zip5 == "Vario" ~ NA,
  zip5 == "UB7 O" ~ NA,
  zip5 == "UB7 " ~ NA,
  .default = zip5
))
#case_when() sets rules for recoding values: here, replacing bad zip codes with NA, otherwise keeping the original.

filter(source == "COMMERCE/MARYLAND TOURISM BOARD") |>
  group_by(zip5) |>
  summarise(total_amount = sum(amount)) |>
  arrange(desc(total_amount))
#Filters rows to only Maryland Tourism Board, groups by zip code, calculates total amounts, and sorts from highest to lowest.
```