---
title: "pre_lab_11.Rmd"
author: "aamaly hossain"
date: "2025-12-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# An intro to text analysis

Throughout this course, we've been focused on finding information in structured data. We've learned a lot of techniques to do that, and we've learned how the creative mixing and matching of those skills can find new insights.

What happens when the insights are in unstructured data? Like a block of text?

Turning unstructured text into data to analyze is a whole course in and of itself -- and one worth taking if you've got the credit hours -- but some simple stuff is in the grasp of basic data analysis.

To do this, we'll need a new library -- [tidytext](https://cran.r-project.org/web/packages/tidytext/vignettes/tidytext.html), which you can guess by the name plays very nicely with the tidyverse. So install it in the console with `install.packages("tidytext")` and we'll get rolling.

### Task 1: Load libraries and settings

**Task** Run the following code in the gray-colored codeblock below to load the libraries we'll use.

```{r}
#install.packages("tidytext")
library(tidyverse)
library(tidytext)
library(janitor)
library(lubridate)
```

Here's the question we're going to go after: what words or phrases appear most in [disciplinary alerts published by the Maryland Board of Physicians](https://www.mbp.state.md.us/disciplinary_2024.aspx)?

To answer this question, we'll use the text of those alerts, which a previous class gathered.

Let's read in this data and examine it:

### Task 2: Read in data

**Task** Run the following code and describe the dataframe it outputs. **Answer** There are 15 columns and 499 rows of data, showcasing Maryland physician cases that have had some type of consequence to their medical practicing license (e.g. revoked, suspended, surrendered, etc.), along with the first and last name (sometimes middle name/inital or suffix), the date and year it happened, the filename and a bit of text describing the matter. 

```{r}
alerts <- read_rds("data/md_doc_alerts.rds")
```

What we want to do is to make the `text` column easier to analyze. Let's say we want to find out the most commonly used words. We'll want to remove URLs from the text of the releases since they aren't actual words. Let's use mutate to make that happen:

### Task 3: Remove URLs from content

**Task** Run the following code.

```{r}
alerts <- alerts |>
  mutate(text = gsub("http.*","", text))
```

If you are trying to create a list of unique words, R will treat differences in capitalization as unique and also will include punctuation by default, even using its `unique` function:

### Task 4: Trying out unique

**Task** Run the following code and describe what the `unique` function does to the original list of words. **Answer** The unique function creates an output only presenting one of the variation of words. In this case, "dog" and "cat" are listed twice as lowercase words, but the unique function categorizes both "dog" and "cat," respectively as one entity. However, it does not categorize "Dog" and "dog" together because the first 'dog' has a capitalization. Additionally, it categorized the "," seperately. In simple terms, this function shortens the list to only show the unique/different variations of the word. 

```{r}
a_list_of_words <- c("Dog", "dog", "dog", "cat", "cat", ",")
unique(a_list_of_words)
```

Fortunately, this is a solved problem with tidytext, which has a function called `unnest_tokens` that will convert the text to lowercase and remove all punctuation. The way that `unnest_tokens` works is that we tell it what we want to call the field we're creating with this breaking apart, then we tell it what we're breaking apart -- what field has all the text in it. For us, that's the `text` column:

### Task 5: Trying out unnest_tokens

**Task** Run the following code and describe the output of using the `unnest_tokens` function. **Answer** The dataframe now has one column and 1,379,101 rows, showcasing the alerts data in all lowercase text and each individual word gets a row without punctuation. 

```{r}
unique_words <- alerts |> select(text) |>
  unnest_tokens(word, text)
View(unique_words)
```

### Task 6: Make a column chart

**Task** Run the following code and describe what the resulting graphic shows. Is it interesting? **Answer** I mean it's not too interesting. I think "the" is a very commonly used word in government and official documents. I do think it's interesting that it was used nearly 100,000. It seems like there are a lot more conjunction words instead of words such as patient or license or Maryland. 

Now we can look at the top words in this dataset. Let's limit ourselves to making a plot of the top 25 words, and we'll use the function `count` to do the counting:

```{r}
unique_words |>
  count(word, sort = TRUE) |>
  top_n(25) |>
  mutate(word = reorder(word, n)) |>
  ggplot(aes(x = word, y = n)) +
  geom_col() +
  xlab(NULL) +
  coord_flip() +
      labs(x = "Count",
      y = "Unique words",
      title = "Count of unique words found in Maryland doctor discipline alerts")
```

Well, that's a bit underwhelming - a lot of very common (and short) words. This also is a solved problem in working with text data, and words like "a" and "the" are known as "stop words". In most cases you'll want to remove them from your analysis since they are so common. Tidytext provides a dataframe of them that we'll load, and then we'll add some of our own.

### Task 7: Load the stop words

**Task** Run the following code

```{r}
data("stop_words")

stop_words <- stop_words |>
  add_row(word = "respondent") |> 
  add_row(word = "board") |> 
  add_row(word = "panel") |>
  add_row(word = "dr") |>
  add_row(word = "maryland") |>
  add_row(word = 'practice') |> 
  add_row(word = '1') |> 
  add_row(word = '2') |>
  add_row(word = '3') |>
  add_row(word = '14')
```

Then we're going to use a function we haven't used yet called an `anti_join`, which filters out any matches. So we'll `anti_join` the stop words and get a list of words that aren't stop words.

From there, we can get a simple word frequency by just grouping them together and counting them. We can borrow the percent code from above to get a percent of the words our top 10 words represent.

### Task 8: Using anti_join

**Task** Run the following code and describe the results. Is it more interesting than before? **Answer** I think it's more interesting now because patient and license are at the top, along with other medical terms. Patient was listed 8,147 times and license followed it at 6,301. Additionally, I think it's interesting that consent was one of the words that were listed in the top ten most common words. 

```{r}
unique_words |>
  anti_join(stop_words) |>
  group_by(word) |>
  tally(sort=TRUE) |>
  mutate(percent = (n/sum(n))*100) |>
  top_n(50)
```

Those seem like more relevant unique words. Now, here's where we can start to do more interesting and meaningful analysis. Let's create three dataframes of unique words based on time: one for all of 2023 and the other for all of 2022:

### Task 9: Create dataframes for 2022 and 2023

**Task** Run the following code

```{r}
unique_words_2023 <- alerts |>
  filter(year(date) == 2023) |>
  select(text) |>
  unnest_tokens(word, text)

unique_words_2022 <- alerts |>
  filter(year(date) == 2022) |>
  select(text) |>
  unnest_tokens(word, text)
```

Then we can create top 10 lists for both of them and compare:

### Task 10: Create dataframes with the top 10 words in each year

**Task** Run the following code and describe the results. **Answer**  The results show that in 2023, the most common meaningful word (after removing stopwords) was "license," making up the highest percentage of all counted words. In 2022, the most frequently used word was "disciplinary." This shows a shift in emphasis between the two years --- 2022 text emphasized disciplinary actions, while 2023 text featured licensing issues more prominently. 

```{r}
unique_words_2023 |>
  anti_join(stop_words) |>
  group_by(word) |>
  tally(sort=TRUE) |>
  mutate(percent = (n/sum(n))*100) |>
  top_n(10)

unique_words_2022 |>
  anti_join(stop_words) |>
  group_by(word) |>
  tally(sort=TRUE) |>
  mutate(percent = (n/sum(n))*100) |>
  top_n(10)
```

Useful, but only to a point. Single words only gets us so much.

## Going beyond a single word

The next step in text analysis is using `ngrams`. An `ngram` is any combination of words that you specify. Two word ngrams are called bigrams (bi-grams). Three would be trigrams. And so forth.

The code to make ngrams is similar to what we did above, but involves some more twists.

So this block is is going to do the following:

1.  Use the alerts data we created above, and filter for pre-2022 releases.
2.  Unnest the tokens again, but instead we're going to create a field called bigram, break apart summary, but we're going to specify the tokens in this case are ngrams of 2.
3.  We're going to make things easier to read and split bigrams into word1 and word2.
4.  We're going to filter out stopwords again, but this time we're going to do it in both word1 and word2 using a slightly different filtering method.
5.  Because of some weirdness in calculating the percentage, we're going to put bigram back together again, now that the stop words are gone.
6.  We'll then group by, count and create a percent just like we did above.
7.  We'll then use top_n to give us the top 10 bigrams.

### Task 11: Create a dataframe with the top 10 two-word phrases for 2021

**Task** Run the following code and describe the results. **Answer** The top two-word phrases for the pre-2022 alerts are dominated by statutory citations and recurring disciplinary terminology. The most frequent bigram is "health occ," referring to the Health Occupations article of Maryland law, followed by "md code" and "code ann." Beyond these legal references, several substantive disciplinary phrases surface as well, including "summary suspension," "medical records," "health care," and "unprofessional conduct." These results show that alerts prior to 2022 consistently focused on statutory authority, emergency actions and issues involving patient care and conduct. 

```{r}
alerts |>
  filter(date < '2024-01-01') |>
  unnest_tokens(bigram, text, token = "ngrams", n = 2) |>
  separate(bigram, c("word1", "word2"), sep = " ") |>
  filter(!word1 %in% stop_words$word) |>
  filter(!word2 %in% stop_words$word) |>
  mutate(bigram = paste(word1, word2, sep=" ")) |>
  group_by(bigram) |>
  tally(sort=TRUE) |>
  mutate(percent = (n/sum(n))*100) |>
  top_n(10)
```

And we already have a different, more nuanced result. We can see that "summary suspension", "unprofessional conduct" and "medical records" are among the top 2-word phrases. What about after 2023?

### Task 12: Create a dataframe with the top 10 two-word phrases for 2023

**Task** Run the following code and describe the results. **Answer** The 2023 results look almost identical to the pre-2022 patterns. The top phrases-again "health occ," "md code," and "code ann" --- dominate the data, indicating ongoing reliance on statue references in the alerts. Similarly, disciplinary and procedural terms such as "summary suspension," "medical records" and "unprofessional conduct" remain prominient. This potentially shows that the themese and languages used in prior years persisted through 2023, with consistent references to medical documentation and professional conduct issues. 

```{r}
alerts |>
  filter(date < '2024-01-01') |>
  unnest_tokens(bigram, text, token = "ngrams", n = 2) |>
  separate(bigram, c("word1", "word2"), sep = " ") |>
  filter(!word1 %in% stop_words$word) |>
  filter(!word2 %in% stop_words$word) |>
  mutate(bigram = paste(word1, word2, sep=" ")) |>
  group_by(bigram) |>
  tally(sort=TRUE) |>
  mutate(percent = (n/sum(n))*100) |>
  top_n(10)
```

And 2024?

### Task 13: Create a dataframe with the top 10 two-word phrases for 2024

**Task** Run the following code and describe the results. **Answer** The 2024 bigram maintains the same general structure but begins to introduce new, more case-specific or procedural phrases. While "health occ" remains the most frequent bigram, terms like "individual 4," "individual 10," and "perhearing conference" newly appear in the top ten. These additions suggest that 2024 alerts include more references to enumerated individuals or steps in formal proceedings, indicating a possible shift toward a more structured, case-oriented format. At the same time, older patterns continue with recurring phrases such as "summary suspension," "unprofessional conduct," and "health care." 

```{r}
alerts |>
  filter(date >= '2024-01-01') |>
  unnest_tokens(bigram, text, token = "ngrams", n = 2) |>
  separate(bigram, c("word1", "word2"), sep = " ") |>
  filter(!word1 %in% stop_words$word) |>
  filter(!word2 %in% stop_words$word) |>
  mutate(bigram = paste(word1, word2, sep=" ")) |>
  group_by(bigram) |>
  tally(sort=TRUE) |>
  mutate(percent = (n/sum(n))*100) |>
  top_n(10)
```

Pretty similar. You'll notice that the percentages are very small; that's not irrelevant but in some cases it's the differences in patterns that's more important.

There are some potential challenges to doing an analysis. For one, there are variations of words that could probably be standardized - maybe using OpenRefine - that would give us cleaner results. There might be some words among our list of stop words that actually are meaningful in this context.

## Sentiment Analysis

Another popular use of text analysis is to measure the sentiment of a word - whether it expresses a positive or negative idea - and tidytext has built-in tools to make that possible. We use word counts like we've already calculated and bring in a dataframe of words (called a lexicon) along with their sentiments using a function called `get_sentiments`. The most common dataframe is called "bing" which has nothing to do with the Microsoft search engine. Let's load it:

### Task 14: Load the bing lexicon and produce sentiments for our 2021-2023 unique words

**Task** Run the following code and describe the results. Do any of the sentiments seem incorrect or counter-intuitive? **Answer** The sentiment analysis using the Bing lexicon shows the most common positive and negative words found in the 2022 and 2023 alerts. in 2023, negative sentiment words such as "violation,” “surrender,” “guilty,” “failed” and “crime” appear frequently, while positive-sentiment words include “patient,” “right,” “appropriate” and “effective.” The 2022 results show a similar pattern: high counts for negative words like “surrender,” “violation,” “complaint” and “failed,” contrasted with positive words such as “patient,” “right,” “appropriate” and “work.” However, several sentiment assignments seem counter-intuitive because words like “patient,” “appropriate” or “complaint” function differently in regulatory or medical contexts than they do in ordinary emotional language. This demonstrates a common limitation of lexicon-based sentiment analysis: it can highlight general trends but may not accurately capture domain-specific meaning or nuances.

```{r}
bing <- get_sentiments("bing")

bing_word_counts_2023 <- unique_words_2023 |>
  inner_join(bing) |>
  count(word, sentiment, sort = TRUE)

bing_word_counts_2022 <- unique_words_2022 |>
  inner_join(bing) |>
  count(word, sentiment, sort = TRUE)

View(bing_word_counts_2023)
View(bing_word_counts_2022)
```

Gauging the sentiment of a word can be heavily dependent on the context, and as with other types of text analysis sometimes larger patterns are more meaningful than individual results. But the potential with text analysis is vast: knowing what words and phrases that public officials employ can be a way to evaluate their actions and priorities. And those words and phrases are data.
