---
title: "Reverse Engineering Project"
author: "Aamaly Hossain"
date: "2025-10-16
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

In this notebook, we are reverse engineering the story, [L.A. is slammed with record costs for legal payouts](https://www.latimes.com/local/lanow/la-me-ln-city-payouts-20180627-story.html)

About: This story investigates how the City of Los Angeles has faced record-breaking legal costs due to lawsuits and settlements across city departments. The Times reporters found that legal payouts -- especially those involving the Los Angeles Police Department -- have surged over the past decade. Using a city databased of more than 18,000 payouts obtained through a public records request, the reporters showed that 2017 had an unprecedented number of multimillion dollar settlements that LAPD-related cases accounted for over 40% of total costs. The story emphasizes how these rising legal costs strain the city's budhet and contribute to financial challenges across department. 

## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}
options(scipen=999) 

library(readr)
library(tidyverse)
library(lubridate)
library(janitor)
library(readxl)
```

## Load and Cleaning Data

In this section, describe the source of the data, write a basic data dictionary for data you are working with, and discuss any caveats or issues you discovered working with this data. 

Answer: 

The data used in this analysis was compiled by the Los Angeles Times through a public records request to the City of Los Angeles. It includes detailed information on every legal payout -- settlements, judgments and other payments made by the city from July 2005 to June 2018. 

Data Sources:
- payouts.xlsx: Master dataset listing each payout, its fiscal year, amount and related case type. 
- casetypes.csv: Defines the type of each case in both how it came from public records and the latimes_type, meaning the clean version which it was categorized (e.g., police misconduct, dangerous conditions, employment dispute).
- departments.csv: Lists city departments involved in each payout and the case each department was attributed to.  
- department-totals.csv: Aggregated payout totals by department. 
- type-totals.csv: Aggregate totals by case type. 

Variables: 
- 'case_number' = unique identifier for each case
- 'amount' = dollar amount of payout (the amount paid for settlement or judgment) 
- 'fiscal_year' = fiscal year where the payout occurred
- 'department' = department associated with the payout
- 'type' = category of legal case (police, traffic, dangerous condition)

Notes and Caveats: 


```{r}
# Load required data

# Path to data should be loaded from folder "data" i.e. read_csv("data/name_of_data.csv")
casetypes <- read_csv("data/casetypes.csv")
departments <- read_csv("data/departments.csv")
payouts <- read_excel("data/payouts.xlsx", skip = 1)
department_totals <- read_csv("data/department-totals.csv")
type_totals <- read_csv("data/type-totals.csv")

# Clean required data and prepare for analysis if needed. 
payouts |>
clean_names()

```

## Sentences to Engineer
In this notebook, we are reverse engineering five sentences from the story.

### Sentence 1

* **Beyond the mammoth cases, a pileup of other large payouts has padded the tab, a Times analysis of city payouts found.**: [Paste in sentence to engineer here]
* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here
payouts <- payouts |> 
clean_names() 

city_payouts <- payouts |>
  mutate(range = case_when(
    amount < 500000 ~ "<$500K",
    amount >= 500000 & amount < 5000000 ~ "$500K–$5M",
    amount >= 5000000 ~ ">$5M"
  )) |>
  group_by(range) |>
  summarise(total_amount = sum(amount))

# Display results of code below this codeblock
city_payouts 
```

### Sentence 2

* **Last budget year, the city paid out 30 settlements of $1 million or more — five times as many as a decade before.**: [Paste in sentence to engineer here]
* **I was able to confirm the finding by filtering all the payouts above $1 million and then grouping by fiscal year. 2017 had 30 payouts over a million as the sentence says. For a decade before (2007), there were 6 payouts, which means 2017 was five times the amount of 2007.**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here
payouts|>
clean_names() |>
filter(amount >= 1000000) |>
  group_by(fiscal_year) |>
  summarise(total_over_million = n()) |>
  arrange(desc(total_over_million))
# Display results of code below this codeblock
30/6
```

### Sentence 3

* **The Times analyzed a city database listing more than 18,000 payouts dating back to July 2005, obtained through a public records request.**: [Paste in sentence to engineer here]
* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here
nrow(payouts)

# Display results of code below this codeblock
#18,703
```

### Sentence 4

* **Thousands of legal battles involving the Los Angeles Police Department … have added up to more than 40% of the total — a higher sum than any other city department, the analysis showed.**: [Paste in sentence to engineer here]
* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here
department_totals |>
mutate(pct_of_total = (amount/sum(amount)) * 100) |>
arrange(desc(pct_of_total))
# Display results of code below this codeblock
#LAPD == 42.43863555145 <- total percent
```

### Sentence 5

* **Legal payouts for ‘dangerous conditions’ have rivaled and, last budget year, even exceeded the cost of lawsuits over police misconduct, the analysis found.**: [Paste in sentence to engineer here]
* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here
payouts <- payouts |>
  mutate(case_type = str_trim(case_type),       
         case_type = str_to_title(case_type))  

dangerous_totals <- payouts |>
  filter(str_detect(case_type, "Dangerous")) |>
  group_by(fiscal_year) |>
  summarise(dangerous_total = sum(amount)) |>
arrange(desc(fiscal_year))

police_totals <- payouts |>
  filter(str_detect(case_type, "Police")) |>
  group_by(fiscal_year) |>
  summarise(police_total = sum(amount)) |>
arrange(desc(fiscal_year))

dangerous_totals
police_totals

# Display results of code below this codeblock
dangerous_vs_police
```

-30-