---
title: "Reverse Engineering Project"
author: "Aamaly Hossain"
date: "2025-10-16
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

In this notebook, we are reverse engineering the story, [L.A. is slammed with record costs for legal payouts](https://www.latimes.com/local/lanow/la-me-ln-city-payouts-20180627-story.html)

## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}
options(scipen=999) 

library(readr)
library(tidyverse)
library(lubridate)
library(janitor)
```

## Load and Cleaning Data

In this section, describe the source of the data, write a basic data dictionary for data you are working with, and discuss any caveats or issues you discovered working with this data. 

```{r}
# Load required data

# Path to data should be loaded from folder "data" i.e. read_csv("data/name_of_data.csv")
casetypes <- read_csv("data/casetypes.csv")
departments <- read_csv("data/departments.csv")
payouts <- read_excel("data/payouts.xlsx", skip = 1)
department_totals <- read_csv("data/department-totals.csv")
type_totals <- read_csv("data/type-totals.csv")

# Clean required data and prepare for analysis if needed. 
payouts |>
clean_names()

```

## Sentences to Engineer
In this notebook, we are reverse engineering five sentences from the story.

### Sentence 1

* **Beyond the mammoth cases, a pileup of other large payouts has padded the tab, a Times analysis of city payouts found.**: [Paste in sentence to engineer here]
* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Sentence 2

* **Last budget year, the city paid out 30 settlements of $1 million or more — five times as many as a decade before.**: [Paste in sentence to engineer here]
* **I was able to confirm the finding by filtering all the payouts above $1 million and then grouping by fiscal year. 2017 had 30 payouts over a million as the sentence says. For a decade before (2007), there were 6 payouts, which means 2017 was five times the amount of 2007.**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here
payouts|>
clean_names() |>
filter(amount >= 1000000) |>
  group_by(fiscal_year) |>
  summarise(total_over_million = n()) |>
  arrange(desc(total_over_million))
# Display results of code below this codeblock
30/6
```

### Sentence 3

* **The Times analyzed a city database listing more than 18,000 payouts dating back to July 2005, obtained through a public records request.**: [Paste in sentence to engineer here]
* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Sentence 4

* **Thousands of legal battles involving the Los Angeles Police Department … have added up to more than 40% of the total — a higher sum than any other city department, the analysis showed.**: [Paste in sentence to engineer here]
* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Sentence 5

* **Legal payouts for ‘dangerous conditions’ have rivaled and, last budget year, even exceeded the cost of lawsuits over police misconduct, the analysis found.**: [Paste in sentence to engineer here]
* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

-30-