---
title: "lab_03"
author: "Aamaly Hossain"
date: "2025-09-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## TASK: Load libraries

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse
library(tidyverse)
# Load janitor
library(janitor)
```

## TASK: Load data and clean up the column names

Load the two datasets we worked with in the pre_lab that are in the data folder: the Maryland payments to vendors and Maryland grants and loans AND, using janitor, change the column names so that they are lower-cased and any spaces are replaced with underscores.

```{r}
# Load in the Maryland state agency payments data, making sure that it correctly sets the datatypes using all of the rows
payments <- read_csv("data/State_of_Maryland_Payments_Data__FY2008_to_FY2024.csv", guess_max=369008)
# Load in the Maryland grants and loan data
md_grants_loans <- read_csv("data/State_of_Maryland_Grant_and_Loan_Data__FY2009_to_FY2022.csv", guess_max = 19482)
```

## Answer questions

Q1: You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed in the payments data. Look at & search the records in the dataframe to start your search, starting with a broad search and then narrowing it down. Then write code to find rows where the vendor name contains some version of the county's school system.

HINT: you should use filter() just once, and inside your filter you want to look for the variations one at a time. Think "this version OR that version."

When you have found as many as possible, save the result to a new dataframe called `pg_schools`. How many records did you find? 

A1: I found that there were 70 rows -- where an agency has paid Prince George's County Public Schools 

```{r}
cleaned_payments <- payments |>
  clean_names()

cleaned_payments |>
  mutate(agency_name = str_to_upper(agency_name), vendor_name = str_to_upper(vendor_name)) |>
  distinct() 
  

pg_schools <- cleaned_payments|>
  filter(
    str_detect(vendor_name, "PRINCE GEORGE") & 
    str_detect(vendor_name, "PUBLIC SCHOOL") |
    str_detect(vendor_name, "PRINCE GEORGES COUNTY SCHOOL") |
    str_detect(vendor_name, "PG COUNTY PUBLIC SCHOOL") |
    str_detect(vendor_name, "PGCPS") |
    str_detect(vendor_name, "PG COUNTY PUBLIC SCHOOL") |
    str_detect(vendor_name, "PG PUBLIC SCHOOL") |
    str_detect(vendor_name, "PRINCE GOERGE PUBLIC SCHOOL") |
    str_detect(vendor_name, "PRINCE GOERGE COUNTY PUBLIC SCHOOL") |
    str_detect(vendor_name, "PG PUBLIC SCHOOL"))
view(pg_schools)
```

Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? Are there any aspects of the results that stand out to you?

A2: The Maryland State Department of Education has the largest total amount in 2021 -- 1,401,301,255.45 dollars given to the Prince George's County school system (very large number -- more than $1 billion). I think this makes sense because public schools I've reported for usually recieve that much; however, I do think it's interesting that this year was the highest -- possibly due to virtual learning that year because of COVID (schools probably had to distribute several laptops and bring in new appliances and other school related material). Additionally, the fact that the first 11 rows, that Prince George's County Public Schools, get paid from are all the Maryland State Department of Education is very interesting and stands out within the results. However, again, it makes sense because the state education department needs to fund the public schools throughout Prince George's County. It's also crucial to note that the public school construction didn't rank higher in the list than MSDE, which I would assume it did.  

```{r}
pg_schools |>
  group_by(agency_name, fiscal_year) |>
  summarise(
    total_amount = sum(amount)
  ) |>
  arrange(desc(total_amount))
```

Q3. You're interested in examining the vendors paid for capital projects at the University of Maryland. Search the payments data to find those payments, then write code to produce a list of each vendor and the total amount it received.

Look carefully at that list - what cleaning will it require to ensure that you have only one row per vendor and zip code? Provide at least three examples of vendor names that are similar and should be combined into one. Describe in words how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?

A3: Three examples of vendor names that are similar: 
    1. "WHITING TURNER CONTRACATING CO" and "WHITING TURNER CONTRACTING COMPANY"
    2. "AYERS SAINT GROSS" and "AYERS SAINT GROSS INCORPORATED" and "AYERS SAINT GROSS INC" 
    3. "BALLINGER" and "BALLINGER CO" 
    I think first, I would remove punctuation like commas, periods, dashes, and then standardize common abbreviations using AP Style to have the full CO, INC, CORP or LLC in the vendor_name. Considering all my cleaned_payments are already str_detect to be uppercase, that would prevent mismatches. Also, I would make sure there are no extra spaces in the vendor_name, and ensure there's a unified zip code. I think I would just use a filtering code, str_detect and mutate in order to do this and replace all the punctuation, clean the names using case_when and mutate. 

```{r}
umd_vendors <- cleaned_payments |>
  filter(
    str_detect(agency_name, fixed("CAPITAL PROJECTS- UNIVERSITY OF MARYLAND")) 
  ) |>
  group_by(vendor_name, vendor_zip, agency_name) |>
  summarise(total_amount = sum(amount, na.rm = TRUE)) |>
  arrange(desc(total_amount))

view(umd_vendors)

```

Q4: You want to understand how much state grant money went to stem cell research projects each fiscal year. Search the data to find grants that fit that description. Once you find which column(s) contain the information you want, your code should find just the grants for stem cell research in a single filter. Then calculate the total amount of money for each fiscal year. Write a sentence summarizing how stem cell research funding has changed over time.

A4: Stem cell research funding has fluctuated over the years, peaking in 2009, declining through much of the 2010s, and then rising sharply in 2023 to its highest level since 2009. 

```{r}
md_grants_loans <- read_csv("data/State_of_Maryland_Grant_and_Loan_Data__FY2009_to_FY2022.csv", guess_max = 19482)

cleaned_md_grants_loans <- md_grants_loans |>
  clean_names()

cleaned_md_grants_loans <- cleaned_md_grants_loans |>
  mutate(
    grantor = str_to_upper(grantor), 
    grantee = str_to_upper(grantee), 
    description = str_to_upper(description))|>
  distinct() 

stem_cell_year_amount <- cleaned_md_grants_loans |>
  filter(
    str_detect(description, "STEM CELL") &
      str_detect(description, "RESEARCH")) |>
  group_by(fiscal_year) |>
  summarise(
    total_amount = sum(amount)
  ) |>
  arrange(desc(fiscal_year))

view(stem_cell_year_amount)
```

Q5: Using the Maryland grants and loan data, find records where the zip code is the one where the University of Maryland-College Park is located in and produce a list of grantees from that zip code, along with the total amount they received. Write a paragraph explaining how many grantees are there, what you would need to do to clean the results so that you had one record per grantee and which grantees are most interesting to you?

A5: From the cleaned_md_grants_loans for the zip code of University of Maryland, College Park, there are 18 grantees listed, but several entries appear to be different names for that same institution, which indicates duplicates, also known as dupes! To clean the results and have one record per grantee, I would need to standardize the institution names -- that just repeat different ways to indicate University of Maryland, College Park -- and sum its total amounts. So, would probably be using mutate. Once cleaned, the list would more accurately reflect unique recipients. The most interesting grantee include the main University of Maryland entities, which together receive millions in funding, as well as smaller innovative companies like Zymetis Inc, Remedium Technologies Inc and Leukosight Inc, which could mean there's emerging research initiatives at UMD or the College Park area. 

```{r}
#University of Maryland zip code: 20742

cp_grants_loans <- cleaned_md_grants_loans |>
  filter(str_detect(zip_code, "20742")) |>
  group_by(grantee) |>
  summarise(
    total_amount = sum(amount)
  ) |>
  arrange(desc(total_amount))

view(cp_grants_loans)
```
