---
title: "pre_lab_08.Rmd"
author: "Aamaly Hossain"
date: "2025-11-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Chapter 26

Visualizing data is becoming a much greater part of journalism. Large news organizations are creating graphics desks that create complex visuals with data to inform the public about important events.

To do it well is a course on its own. And not every story needs a feat of programming and art. Sometimes, you can help yourself and your story by just creating a quick chart, which helps you see patterns in the data that wouldn't otherwise surface.

Good news: one of the best libraries for visualizing data is in the tidyverse and it's pretty simple to make simple charts quickly with just a little bit of code. It's called [ggplot2](https://ggplot2.tidyverse.org/).

Let's revisit some data we've used in the past and turn it into charts. First, let's load libraries. When we load the tidyverse, we get ggplot2.

### Task 1: Load packages

**Task** Run the following code to load packages.

```{r}
library(tidyverse)
library(lubridate)
```

The dataset we'll use is 911 overdose calls from Baltimore County in 2022. Let's load it.

### Task 2: Load data

**Task** Run the following code to load the data. In the space below, indicate how many rows and columns it has. **Answer** The data has 3822 rows and 4 columns. 

```{r}
baltco_911_calls <- read_csv("data/baltco_911_calls.csv")
```

## Bar charts

The first kind of chart we'll create is a simple bar chart.

It's a chart designed to show differences between things -- the magnitude of one thing, compared to the next thing, and the next, and the next.

So if we have thing, like a county, or a month, or a day of the week, and then a count of that group, we can make a bar chart.

So what does the chart of the top months with the most 911 overdose calls look like?

First, we'll add a month column to our dataframe using lubridate.

### Task 3: Add a month column to our 911 calls dataframe using lubridate and calculate the number of calls for each month

**Task** Run the following code to add a month column to our dataframe using lubridate and count the number of calls for each month. In the space below, explain what each step of the code below is doing in English. Note the "label=TRUE" part in the mutate statement. **Answer** The first line of code shows the original Baltimore 911 calls datafram being called a new dataframe to signify that months were added. Then, the second line of code shows the mutate function adding a new column called months to the dataframe. The month function from the lubridate package extracts the month from the date column in the original dataframe to put it in the new dataframe. The argument "label=TRUE" makes the month appear as a name instead of a number. Then, the third line of code groups all the 911 call records by month so we can perform calculation, like counting, within each month. The summarize function creates a new summarized table, while the "n" function counts the number of rows (calls) in each group, so the total calls becomes the number of 911 calls per month. And this is then ordered with the arrange function, resulting in the most to fewest calls. And then the new dataframe by itself at the bottom displays the results, showing each month and the total number of 911 calls that occurred in each month. 

```{r}
baltco_911_calls_by_month <- baltco_911_calls |>
  mutate(month = month(date, label=TRUE)) |> 
  group_by(month) |> 
  summarize(total_calls = n()) |> 
  arrange(desc(total_calls))

baltco_911_calls_by_month
```

Now let's create a bar chart using ggplot.

With ggplot, the first thing we'll always do is draw a blank canvas that will contain our chart. We start with our dataframe name, and then (\|\>) we invoke the ggplot() function to make that blank canvas. All this does is make a gray box, the blank canvas that will hold our chart.

### Task 4: Create an empty ggplot container

**Task** Run the following code. Explain what you see after running the code in the space below. **Answer** I see a gray box -- a blank canvas, which is what it should be. 

```{r}
baltco_911_calls_by_month |>
  ggplot()
```

Next we need to tell ggplot what kind of chart to make.

In ggplot, we work with two key concepts called geometries (abbreviated frequently as geom) and aesthetics (abbreviated as aes).

Geometries are the shape that the data will take; think of line charts, bar charts, scatterplots, histograms, pie charts and other common graphics forms.

Aesthetics help ggplot know what component of our data to visualize -- why we'll visualize values from one column instead of another.

In a bar chart, we first pass in the data to the geometry, then set the aesthetic.

In the codeblock below, we've added a new function, geom_bar().

Using geom_bar() -- as opposed to geom_line() -- says we're making a bar chart.

Inside of that function, the aesthetic, aes, says which columns to use in drawing the chart.

We're setting the values on the x axis (horizontal) to be the abbreviation of the month. We set weight to total calls, and it uses that value to "weight" or set the height of each bar.

One quirk here with ggplot.

After we've invoked the ggplot() function, you'll notice we're using a + symbol. It means the same thing as \|\> -- "and then do this". It's just a quirk of ggplot() that after you invoke the ggplot() function, you use + instead of \|\>. It makes no sense to me either, just something to live with.

### Task 5: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below. **Answer** I see a bar graph, displaying the month on the x-axis and the number of calls by hundreds in the y-axis, with gray bars signifying the number of calls for each month. The months are abbreviated for space and aesthetic I believe, and the y-axis starts at 0. There is no title. The months are categorized in chronological order (correct month order). 

```{r}
baltco_911_calls_by_month |>
  ggplot() +
  geom_bar(aes(x=month, weight=total_calls))
```

This is a very basic chart. But it's hard to derive much meaning from this chart, because the months aren't ordered from highest to lowest by total_calls We can fix that by using the reorder() function to do just that:

### Task 6: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below. What has changed from the last codeblock? **Answer** Now, the bars are ascending by calls. So, the least number of calls is shown as the first bar in the graph, which is the month of February. It's not in chronological order, but in ascending order of the 911 calls in each month -- least to greatest. Essentailly, the x-axis was reordered. 

```{r}
baltco_911_calls_by_month |>
  ggplot() +
  geom_bar(aes(x=reorder(month,total_calls), weight=total_calls))
```

This is a little more useful. But the bottom is kind of a mess, with overlapping names. We can fix that by flipping it from a vertical bar chart (also called a column chart) to a horizontal one. coord_flip() does that for you.

### Task 7: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below. What has changed from the last codeblock? **Answer** Now, the x-axis is vertical instead of horizantal, vice versa with the y-axis. Instead, of vertical bars there are horizontal bars, which also ascend from the bottom-up. And you can clearly see the reordered months in correlation to ascending and descending pattern too. Essentailly, the coordinates flipped. 

```{r}
baltco_911_calls_by_month |>
  ggplot() +
  geom_bar(aes(x=reorder(month,total_calls), weight=total_calls)) +
  coord_flip()
```

Is this art? No. Does it quickly tell you something meaningful? It does.

We're mainly going to use these charts to help us in reporting, so style isn't that important.

But it's worth mentioning that we can pretty up these charts for publication, if we wanted to, with some more code. To style the chart, we can change or even modify the "theme", a kind of skin that makes the chart look better.

It's kind of like applying CSS to html. Here I'm changing the theme slightly to remove the gray background with one of ggplot's built in themes, theme_minimal().

### Task 8: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below. What has changed from the last codeblock? **Answer** There isn't a grey background underneath the bars. You can see from the codeblock that there has been an additional line of code called theme_minimal(), showcasing why the grey background isn't there, but now it's a more minimalist bar graph. 

```{r}
baltco_911_calls_by_month |>
  ggplot() +
  geom_bar(aes(x=reorder(month,total_calls), weight=total_calls)) +
  coord_flip() + 
  theme_minimal()
```

The ggplot universe is pretty big, and lots of people have made and released cool themes for you to use. Want to make your graphics look kind of like [The Economist's](https://www.economist.com/) graphics? There's a theme for that.

First, you have to install and load a package that contains lots of extra themes, called [ggthemes](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/).

### Task 9: Run the following code

**Task** Run the following code.

```{r}
install.packages('ggthemes')
library(ggthemes)
```

And now we'll apply the economist theme from that package with theme_economist()

### Task 10: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below. What has changed from the last version of the chart? **Answer** I think the flipped x-axis is way too close to the actual months. Now, the theme is economist, so I guess that means a blue-tinted background and a darker color for the font. And it seems like it's a lot closer together than the last chart. 

```{r}
baltco_911_calls_by_month |>
  ggplot() +
  geom_bar(aes(x=reorder(month,total_calls), weight=total_calls)) +
  coord_flip() + 
  theme_economist()
```

Those axis titles are kind of a mess. Let's change "count" on the x axis to "total calls" and change "reorder(month,total_calls)" to "month". And while we're at it, let's add a basic title and a source as a caption. We'll use a new function, labs(), which is short for labels.

### Task 11: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below. What has changed from the last codeblock? **Answer** Now, there's a title, a renamed x-axis and y-axis and a caption to the source. However, the economist theme is stil the same. The x-axis and y-axis titles are very close to the units.  

```{r}
baltco_911_calls_by_month |>
  ggplot() +
  geom_bar(aes(x=reorder(month,total_calls), weight=total_calls)) +
  coord_flip() + 
  theme_economist() +
  labs(
    title="More 911 Overdose Calls in Warmer Months",
    x = "month",
    y = "total calls",
    caption = "source: Baltimore County"
    
  )
```

Viola. Not super pretty, but good enough to show an editor to help them understand the conclusions you reached with your data analysis.

## Line charts

Let's look at how to make another common chart type that will help you understand patterns in your data.

Line charts can show change over time. It works much the same as a bar chart, code wise, but instead of a weight, it uses a y.

So, let's use the same Baltimore County data and generate a count of contributions for each date in our data.

### Task 12: Create a dataframe of 911 calls by date

**Task** Run the following code. Explain what you see after running the code in the space below. **Answer** I see the total number of 911 calls in every day of February 2022, in no particular order. 

```{r}
baltco_911_calls_by_date <- baltco_911_calls |>
  group_by(date) |>
  summarise(
    total_calls=n()
  )

baltco_911_calls_by_date 
```

And now let's make a line chart to look for patterns in this data.

We'll put the date on the x axis and total contributions on the y axis.

### Task 13: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below. **Answer** I see a black, lined pattern that spikes up and down, showing the progression of calls from February/March 2022 to the end of Jan 2023. I see the x-axis is the dates and the total_calls is the y-axis. There is no title. Nonetheless, there are drastic spikes in July, early October and January. It's not super easy to read, but I think it makes a good way to display the outliers -- or what could make a good story. Additionally, the number of calls usually ranges between 5-20, with everything else outside of the range is an outlier. 

```{r}
baltco_911_calls_by_date |>
  ggplot() + 
  geom_line(aes(x=date, y=total_calls))
```

It's not super pretty, but there's a bit of a pattern here: the number of calls fluctuates between 5 and 20 a day for most of this period, and then jumps way up at certain points during the year. In particular there are spikes in July, early October and January.

Right now, it's kind of hard to see specifics, though. When did some of those smaller spikes and troughs happen?

We can't really tell. So let's modify the x axis to have one tick mark and label per month. We can do that with a function called scale_x_date().

We'll set the date_breaks to appear for every week; if we wanted every month, we'd say date_breaks = "1 month". We can set the date to appear as month abbreviated name (%b) and day (%d).

### Task 14: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below. **Answer** Oh wow -- it seems like the x-axis is all jumbled up, but there has been an addition of date-breaks as seen on the x-axis and the background of the main line graph. The date breaks have been spread 1 week apart. 
```{r}
baltco_911_calls_by_date |>
  ggplot() + 
  geom_line(aes(x=date, y=total_calls)) + 
  scale_x_date(date_breaks = "1 week", date_labels = "%b %d")
```

Those are a little hard to read, so we can turn them 45 degrees to remove the overlap using the theme() function for styling. With "axis.text.x = element_text(angle = 45, hjust=1)" we're saying, turn the date labels 45 degrees.

### Task 15: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below. **Answer** The x-axis dates have been turned to a 45 degree angle, removing the overlap that made the 1 week apart dates hard to read. it's not the best formatted because it's not easy to read, but it's doable and good for starters. And there are a lot of outliers, but the range seems to be in the 5 to 20 calls. And the dates range from Jan 2022 to Feb 2023. 

```{r}
baltco_911_calls_by_date |>
  ggplot() + 
  geom_line(aes(x=date, y=total_calls)) + 
  scale_x_date(date_breaks = "1 week", date_labels = "%b %d") +
  theme(
    axis.text.x = element_text(angle = 45,  hjust=1)
  )
```

Again, this isn't as pretty as we could make it. But by charting this, we can quickly see some outliers that could guide our reporting.

We're just scratching the surface of what ggplot can do, and chart types. There's so much more you can do, so many other chart types you can make. But the basics we've shown here will get you started.
