---
title: "lab_11"
author: "Aamaly Hossain"
date: "2025-12-10"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   Our usual libraries for working with data, plus rvest and lubridate.

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.
```{r}
#install.packages("tidytext")
library(tidyverse)
library(tidytext)
library(janitor)
library(lubridate)
library(stringr)
library(rvest)
library(dplyr)
```

Let's get to scraping.

## Questions

**Q1**. Scrape the listing of available Maryland physician disciplinary alerts at <https://www.mbp.state.md.us/disciplinary_2024.aspx> into a dataframe. You should have three columns, one of which is a date, so make sure the date column has a date datatype. What's the most common sanction, and how many times is it mentioned in the 2024 alerts?

**A1** Based on the dataframe, and ensuring that each sanction that is listed differently is its own unique sanction, the "Summary Suspension" sanction was the most common in 2024 alerts. It was mentioned eight times. This was not the calculation of how many times suspension was listed in 24 alerts. But, if we were to look at how many times it was mentioned in 2024 alerts, specifically the word "suspension," it was mentioned 33 times in 2024 alerts. However, "Summary Suspension" was the most common alert as a unique sanction. And "suspension" was mentioned 33 times. 

```{r}
url <- "https://www.mbp.state.md.us/disciplinary_2024.aspx"

page <- read_html(url)

alerts_raw <- page |>
  html_element("table") |>
  html_table() |>
  clean_names()

alerts_raw

alerts <- alerts_raw |>
  mutate(date = floor_date(mdy(date), unit = "day"),
         sanction = str_to_upper(str_remove_all(sanction, "[[:punct:]]")))
alerts

sanction_counts <- alerts |>
  count(sanction, sort = TRUE)

sanction_counts

counts <- alerts |>
  filter(str_detect(sanction, "SUSPENSION")) |>
  nrow()

counts
```

**Q2** Next, let's scrape the list of press releases from Maryland's Office of the Public Defender, <https://www.opd.state.md.us/press-releases>. This isn't a table, so you'll need to use `html_elements()` and your browser's inspector and do some clean up on the results. The result should be a dataframe with two columns that contain the date and title, and the date column should have a date datatype. The challenge here is figuring out how to isolate the releases.

When you finish scraping into a dataframe, write code to find the press releases that have the word "police" in the title.

**A2** There are nine rows of data, meaning nine press releases that have the word "police" in the title. Some of these press releases have the word "police" repeatedly in the title. 

```{r}
url <- "https://www.opd.state.md.us/press-releases"


opd <- read_html(url)


raw <- opd |>
  html_elements("div[data-testid='richTextElement'] *") |>
  html_text2() 

raw <- str_trim(raw)

raw_with_dates <- keep(raw, ~ str_detect(.x, "\\b\\d{1,2}/\\d{1,2}/\\d{4}\\b"))

press_df <- tibble(raw = raw_with_dates) |>
  distinct(raw, .keep_all = TRUE) |>
  mutate(
    date_str = str_extract(raw, "\\d{1,2}/\\d{1,2}/\\d{4}"),
    date = mdy(date_str),
    title = str_remove(raw, paste0("^", date_str, "\\s*[:\\-]?\\s*")),
    title = str_trim(title)
  ) |>
  filter(!is.na(date) & title != "")

press_df <- press_df |> select(date, title)

press_df

police_df <- press_df |>
  filter(str_detect(str_to_lower(title), "police"))

police_df


```

**Q3** Sen. Chris Van Hollen, D-Maryland, has posted press releases at <https://www.vanhollen.senate.gov/news/press-releases>. It would be great to have them in a dataframe that has the following columns: date, title and url.

To do this, you will need to scrape the page's html and save that to a variable, and *then* extract the dates, titles and urls into separate dataframes using html_elements(). There are two different elements that contain the press release URL; pick one of them, and you may need to invoke an HTML `class` value to do it. The function `html_text()` pulls out the contents of a tag, but for urls we want the HTML attribute. Rvest gives you a way to extract the URL from a link; google to find out what it is. Remember how we turn a list into a dataframe.

At the end, you'll have three dataframes that you want to combine into a single dataframe. When we want to combine the rows of identical dataframes, we used `bind_rows()`. If you were combining columns instead of rows, there's a similar function. Find out what it is and use it to put all of the dataframes toge ther into a single one.

When you're done, rename the columns so they make sense, then make sure the date column is an actual date.

Finally, tell me what questions you could ask of this data if you had all of his press releases. Be creative.

**A3** 

1. Which topics or keywords appear most frequently in the press release titles, and how has this changed over time?

2. How many press releases are published each year or each month—are there seasonal trends or spikes around elections?

3. Are there clusters of press releases on the same day or week, suggesting coordinated messaging?

4. Which press releases mention specific legislation, committees, or agencies most often?

5. Can you identify shifts in tone or focus in the titles over time (e.g., more urgent, positive, or neutral messaging)?

Had to Google a lot of this because I was getting several errors when running my code or I was not getting the correct dataframe. 

```{r}
base_url <- "https://www.vanhollen.senate.gov"


url <- paste0(base_url, "/news/press-releases")
page <- read_html(url)


items <- page |> html_elements("li")


press_items <- items |> keep(~ {
  txt <- html_text2(.x)
  str_detect(txt, "^\\s*\\d{2}\\.\\d{2}\\.\\d{2}")
})


press_df <- press_items |> map_df(function(item) {

  full_text <- html_text2(item)
  date_str <- str_extract(full_text, "^\\s*(\\d{2}\\.\\d{2}\\.\\d{2})") |> str_trim()
  date <- mdy(str_replace_all(date_str, "\\.", "/")) 


  a <- item |> html_element("a")
  

  if (is.null(a)) return(NULL)
  
  title <- a |> html_text2() |> str_squish()
  href <- a |> html_attr("href")
  url <- paste0(base_url, href)

  tibble(date = date, title = title, url = url)
})

press_df
```


```

