---
title: "pre_lab_03.Rmd"
author: "Aamaly Hossain"
date: "2025-09-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
```

## About this notebook

This notebook contains code and explanatory text that your should review and run as you read through two chapters on data cleaning from the course textbook, "Data Journalism with R and the Tidyverse". Answer questions and edit the document as directed.

Running this notebook will help you understand key data analysis methods and concepts that you will put into practice during this week's lab.

When you are finished running the code in this notebook, you will push changes to your course GitHub repo, and upload the link to ELMS as instructed.

## Data Cleaning, Part I

### Task 1: Load libraries and settings

**Task** Run the following code in the gray-colored codeblock below to load the tidyverse library and turn off scientific notation.

```{r}
# Remove scientific notation
#options(scipen=999)
# Load the tidyverse   
library(tidyverse)
```

### Task 2: Load data

**Task** Load some [Maryland state government payments data](https://opendata.maryland.gov/Budget/State-of-Maryland-Payments-Data-FY2008-to-FY2024/7syw-q4cy) by running the following code. We'll use the guess_max() function as an argument to use the first 10 rows to set the data type. What does the first line of the red Warning message that prints out when you load the data say? What happens when you follow its instructions? Answer below. **Answer** It says, "Warning: One or more parsing issues, call 'problems()' on your data frame for details, e.g.: 
dat <- vroom(...)
problems(dat)" 

It produces a table of all the parsing problems. It has 369 rows, which means there are some problems, but not a huge number since the original dataset has 369,000 rows. Based on the reading, the readr library guessed that a given column was of a "double" data type -- a number. Where there was a character string or looked like a date -- it generated new information and guessed the 10 rows. 


```{r}
payments <- read_csv("data/State_of_Maryland_Payments_Data__FY2008_to_FY2024_20250115.csv", guess_max=10)

problems(payments)
```

### Task 3: Reload data

**Task** Run the following codeblock to reload the data, using every row to set the data types. Does it show any parsing errors when you run? Answer below **Answer** No! It doesn't show any parsing errors this time. We can see how many rows and columns there are. 

```{r}
payments <- read_csv("data/State_of_Maryland_Payments_Data__FY2008_to_FY2024_20250115.csv", guess_max=369008)
```

### Task 4: Examine the data with glimpse

**Task** Run the following codeblock to glimpse the data. What data type is the "Vendor Name" field? What data type is the "Amount" field? What data type is the "Date" column? Answer below. **Answer** The "Vendor Name" field is a character, the "Amount" field is a number and the "Date" column is a character. 

```{r}
glimpse(payments)
```

Things that should be characters -- like agency name, vendor name -- are characters (chr). Things that should be numbers (dbl) -- like amount and fiscal year -- are numbers. We've seen before that sometimes dates aren't defined as date datatypes by R - we can fix that using `lubridate`.

### Task 5: Detect wrong spatial data

The second smell we can find in code is **wrong spatial data**. Spatial data means data that refers to some geography; in this dataset the only geographical element is the vendor's zip code. Zip codes should be, at a minimum, five characters long (although composed of numbers, zip codes aren't used as numbers).

We can check to see if any of the zip codes are less than five characters by using [a function called `str_length`](https://stringr.tidyverse.org/reference/str_length.html) inside a filter.

**Task** Run the following codeblock to group by the vendor's zip code and show any that are less than five characters in length. How many zip codes do not have five characters? What do you think explains those records? **Answer** There are 553 rows of unique vendor zip that doesn't have five characters in the zip code. I think this dataset is also showing the number of times each zip code was shown, which would increase the number of zip codes, but they wouldn't be unique. I think this shows that either this is missing a leading zero or some other character.  

```{r}
payments |>
  group_by(`Vendor Zip`) |>
  filter(str_length(`Vendor Zip`) < 5) |> 
  summarise(
    count=n()
  ) |>
  arrange(desc(count))
```

### Task 6: Load Maryland grant and loan data

Let's now look at **gaps in data**. These often occur when you have a date or time element in your data, but there are other potential gaps, too. To illustrate those, we're going to introduce some Maryland state grant and loan data from 2009 forward. Let's load it and take a look:

**Task** Run the following codeblock to load Maryland state government grant and loan data

```{r}
md_grants_loans <- read_csv("data/State_of_Maryland_Grant_and_Loan_Data__FY2009_to_FY2022_20250115.csv")
```

Each row represents a recipient of state grant or loan, along with information about their location and the state agency that provided the money. When we talk about gaps, often they indicate the administrative rules. Here's an example: let's count the number of payments in each category (Grant or Loan) by year in this dataset.

### Task 7: Looking at categories

**Task** Run the following codeblock and describe below what each line is doing. Looking at the results, what jumps out at you as potentially problematic? **Answer** The first column is ascending by fiscal year, starting from 2009 until 2023. The category column shows the recipients who received a state grant, a state loan, contract or N/A within that specific year. And then it is grouped by the fiscal year and the category in order to count the rows that received a particular category. I think something that might be potentially problematic is the N/A within the category column.

```{r}
md_grants_loans |> 
  group_by(`Fiscal Year`, Category) |> 
  summarize(count = n()) |> 
  arrange(`Fiscal Year`)
```

Any time you are going to focus on a column for analysis, you should check for unusual values. Are there any unusually large values or unusually small values? Are there any values that raise immediate questions about the data? Let's look at the smallest amounts in the grants and loan data.

### Task 8: Find outliers

**Task** Run the following code to arrange the grants & loan data so that the smallest amount is first. Describe what you see. **Answer** I see the entire dataset for the grants and loans, which is arranged so the smallest amount and all of its row is in the head of the dataset. It seems to be that most of these rows of data show grants within the first fiscal period in a several years. Everything is arranged based on the amount given in ascending order. 

```{r}
md_grants_loans |> 
  arrange(Amount)
```

## Data Cleaning, Part II

### Task 1: Install janitor

**Task** Run the following codeblock to install the janitor package.

```{r}
install.packages('janitor')
```

### Task 2: Load janitor

**Task** Run the following code to load janitor.

```{r}
library(janitor)
```

Let's continue with our Maryland grants and loans data that we worked with in the previous chapter. There are a number of issues with this data set that might get in the way of asking questions and receiving accurate answers. They are:

-   The column names have spaces in them. This isn't a deal-breaker, as we used this dataframe previously. But it does require that you do some things differently when writing code, and ideally you don't want spaces in your column names.
-   Inconsistent capitalization across multiple columns. Sometimes the grantee is capitalized, and other times not. Portions of the grantor name are sometimes capitalized. This issue will ruin your ability to count and add things using those columns.
-   The zip field mixes five digit ZIP codes and nine digit ZIP codes, and some of the records include spaces. If we wanted to group and count the number of loans in a given ZIP code, this inconsistency would not let us do that correctly.
-   The category column is inconsistent and has some missing values.

Let's get cleaning. Our goal will be to build up one block of code that does all the necessary cleaning in order to answer this question: which zip code has gotten the most amount of money from the Maryland Tourism Board?

### Task 3: Use clean_names()

**Task** Run the following codeblock to use the `clean_names()` function from janitor to standardize column names. How does it change the headers? Answer below. **Answer** All the headers for the columns are now lowercase. And there are underscores when needed to separate two words. 

```{r}
# cleaning function
cleaned_md_grants_loans <- md_grants_loans |>
  clean_names()

# display the cleaned dataset
cleaned_md_grants_loans
```

### Task 4: Use rename()

**Task** Run the following codeblock to use the clean_names() function from janitor to standardize column names and then use rename() to change the "grantor" column to "source". With `rename()` the new name comes first!

```{r}
# cleaning function
cleaned_md_grants_loans <- md_grants_loans |>
  clean_names() |> 
  rename(source = grantor)

# display the cleaned dataset
cleaned_md_grants_loans
```

Right now the `source`, `grantee` and `description` columns have inconsistent capitalization. We can fix that using a mutate statement and a function that changes the case of text called `str_to_upper()`. We'll use the same columns, overwriting what's in there since all we're doing is changing case.

### Task 5: Using str_to_upper() to standardize case

**Task** Run the following codeblock to use str_to_upper() to make the source, grantee and description columns upper-cased. Describe what each line is doing. **Answer** The codeblock is first ensuring that the cleaned dataset from the md_grants_loans dataset is being used. Then, cleaning the column heads, renaming the column "grantor" to "source," and finally, utilizing the mutate function to make the source, grantee and description columns to have the same style of case. So, all of the letters are uppercase in those columns, so the data can be read better when coding. And lastly, the cleaned data will be displayed as seen as the very last line of code. 

```{r}
# cleaning function
cleaned_md_grants_loans <- md_grants_loans |>
  clean_names() |> 
  rename(source = grantor) |> 
  mutate(source = str_to_upper(source), grantee = str_to_upper(grantee), description = str_to_upper(description))

# display the cleaned dataset
cleaned_md_grants_loans
```

### Task 6: Check for duplicate rows

**Task** Run the following codeblock to check for duplicate rows using get_dupes(). How many duplicates are possible? **Answer** There are as many as 58 duplicates, but this is not an exact number. Because we would need to confirm that these are actually duplicates, by fixing the code above and adding the distinct() code to the cleaning script. 

```{r}
cleaned_md_grants_loans |>
  get_dupes()
```

### Task 7: Get rid of duplicate rows

**Task** Run the following codeblock to use distinct() to get rid of duplicate rows. How many rows does the new dataframe have? Answer below. **Answer** Now, the new dataframe has 19,453 rows. 

```{r}
# cleaning function
cleaned_md_grants_loans <- md_grants_loans |>
  clean_names() |> 
  rename(source = grantor) |> 
  mutate(source = str_to_upper(source), grantee = str_to_upper(grantee), description = str_to_upper(description)) |> 
  distinct()

# display the cleaned dataset
cleaned_md_grants_loans
```

### Task 8: Clean up ZIP code

The rest of the problems with this data set all have to do with inconsistent format of values in a few of the columns. To fix these problems, we're going to make use of mutate() in concert with "string functions" -- special functions that allow us to clean up columns stored as character strings. The tidyverse package `stringr` has lots of useful string functions that you can look up!

Let's start by cleaning up the zip field. Remember, some of the rows had a five-digit ZIP code, while others had a nine-digit ZIP code, separated by a hyphen or not.

We're going to write code that tells R to make a new column for our zips, keeping the first five digits on the left, and get rid of anything after that by using `mutate()` in concert with `str_sub()`, from the `stringr` package.

**Task** Run the following codeblock to use str_sub() to convert the ZIP codes that have nine digits to five digits, creating a new column. Look at the difference in the result - what changed? **Answer** Now there is a new column that's named "zip5," but it is under a character instead of a "dbl" or an actual number, but the other zip code was under character as well. Additionally, it generates a nine-digit or just over/under 5 digit zip code to a 5 digit zip code. However, there is one row that still says UB7 O as its zip code. 

```{r}
# cleaning function
cleaned_md_grants_loans <- md_grants_loans |>
  clean_names() |> 
  rename(source = grantor) |> 
  mutate(source = str_to_upper(source), grantee = str_to_upper(grantee), description = str_to_upper(description)) |> 
  distinct() |>
  mutate(zip5 = str_sub(zip_code, start=1L, end=5L))


# display the cleaned dataset
cleaned_md_grants_loans
```

### Task 9: Clean up zip5 field more with case_when()

**Task** Run the following codeblock to use case_when() to change clear non-zip codes into NA values.

```{r}
# cleaning function
cleaned_md_grants_loans <- md_grants_loans |>
  clean_names() |> 
  rename(source = grantor) |> 
  mutate(source = str_to_upper(source), grantee = str_to_upper(grantee), description = str_to_upper(description)) |> 
  distinct() |>
  mutate(zip5 = str_sub(zip_code, start=1L, end=5L)) |>
  mutate(zip5 = case_when(
    zip5 == "Vario" ~ NA,
    zip5 == "UB7 O" ~ NA,
    zip5 == "UB7 " ~ NA,
    .default = zip5
  ))

# display the cleaned dataset
cleaned_md_grants_loans
```

### Task 10: Answer our question!

**Task** Write and run code to answer our original question: which zip code has gotten the most amount of money from the Maryland Tourism Board? Where is that zip code, and does that zip code make sense to you as the top recipient? Write a sentence describing the top result. **Answer** The zip code that has gotten the most amount of money from the Maryland Tourism Board was 21202 with 648,069 dollars. This is located in Downtown Baltimore, including the Inner Harbor area, according to the text. I think this does make sense, as downtown Baltimore being the top recipient from the Maryland Tourism Board because of its population density and tourism attractions for many, being the hub of so many attractions and places. Additionally, about 80% of the population in downtown Baltimore is 18-years-old to 64-years-old, so this makes sense. That exact zip code includes Baltimore City, Baltimore-Columbia-Towson and the Metro area. In that 1.6 square miles, there's a populations of 21,506 people, according to census data in 2023. 

```{r}
cleaned_md_grants_loans <- md_grants_loans |>
  clean_names() |> 
  rename(source = grantor) |> 
  mutate(source = str_to_upper(source), grantee = str_to_upper(grantee), description = str_to_upper(description)) |> 
  distinct() |>
  mutate(zip5 = str_sub(zip_code, start=1L, end=5L)) |>
  mutate(zip5 = case_when(
    zip5 == "Vario" ~ NA,
    zip5 == "UB7 O" ~ NA,
    zip5 == "UB7 " ~ NA,
    .default = zip5
  ))

cleaned_md_grants_loans |>
  filter(source == "COMMERCE/MARYLAND TOURISM BOARD") |>
  group_by(zip5) |>
  summarise(
    total_amount = sum(amount)) |>
  arrange(desc(total_amount))
```
