---
title: "lab_05"
author: "Aamaly Hossain"
date: "2025-10-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   Tabula

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse, plus any other packages you will need to clean data and work with dates.
library(tidyverse)
library(lubridate)
library(janitor)
library(readr)
```

## Get Our PDF

We'll be working with the [911 overdose calls from Baltimore County](https://drive.google.com/file/d/1qkYuojGF_6WKFr5aNQxmewDzcKyOiJFr/view?usp=share_link). You'll want to download it to a place you'll remember (like your Downloads folder, or the labs folder in your repository). The goal is to extract the tables within it, export that to a CSV file, load it into RStudio and ask some questions.

## Extract Data from PDF Using Tabula

Start Tabula, then go to <http://127.0.0.1:8080/> in your browser. Click the "Browse" button and find the PDF file and click "open", and then click the "Import button" in Tabula. This will take a few seconds or longer.

This PDF has a single table spread over multiple pages to extract. We're going to make a single dataframe from this table, exporting it to a CSV file that you will load into R. In Tabula, highlight the table and click the "Preview & Export Extracted Data" button. You may want to play with including or excluding the column headers - YOU SHOULD HAVE FIVE COLUMNS OF DATA.

Save the CSV (it should be called `tabula-Baltimore County; Carey, Samantha log OD.csv` by default) to your lab_05/data folder.

From there, you will need to read in the data, and add or fix headers if necessary. You can choose to include the headers from the PDF in your exported CSV files OR to exclude them and add them when importing. `read_csv` allows us to do this ([and more](https://readr.tidyverse.org/reference/read_delim.html)).

## Load and clean up the data in R

You will need to read in and clean up the data so that it can be used for analysis. By "clean" I mean the column headers should not contain spaces and they should have meaningful names, not "x1" or something similar. How you do that is up to you, but you can use select() with or without the minus sign to include or exclude certain columns. You also can use the `rename` function to, well, rename columns. Importantly, you'll need to ensure that any columns containing a date actually have a date datatype. Our friend `lubridate` can help with this.

```{r}
baltcounty_calls <- read_csv("tabula-Baltimore County; Carey, Samantha log OD.csv", col_names = FALSE
   ) |>
  clean_names() |>
  rename(
    date = x1,
    time = x2,
    case_number = x3,
    evtyp = x4,
    location = x5
  ) |>
  mutate(date = mdy(date))

baltcounty_calls
  
```

## Answer questions

Q1. Write code to generate the number of calls that occurred on each date. Which date in 2022 had the most overdose calls, and how many? Look at the total number of rows in your result and explore the range of dates - based on your result, do you believe there are any days with no overdose calls at all? Explain why or why not.

A1. On July 14, 2022 and October 4, 2022 there were 23 overdose calls. So, both dates were tied for most. It seems like the month of January of 2022 is not in the dataset of those overdoses. However, it does begin with February 2022 and ends in February 2023. So, we do see January 2023 overdose calls. I think there could be a possibility that there are some days with no overdose calls, considering February 6, 2023 and June 11, 2022 only had one overdose call. This leads me to believe it could be possible. Also, there are 362 rows. However, January and February of 2023 are added into this mix. With some quick math (not that quick), it shows us that there were 393 days between February 2022 and February 2023, so this means there were 31 days that either were not calculated for or had no overdose calls reported. 

```{r}
calls_per_date <- baltcounty_calls |>
  group_by(date) |>
  summarise(
    num_calls = n()
  ) |>
  arrange(desc(num_calls))

calls_per_date

summary_info <- calls_per_date |>
  summarise(
    total_days = n(),
    first_date = min(date),
    last_date = max(date),
  )

summary_info
```

Q2. You want to understand if there's a pattern in the day of the week that overdose calls are made. Add a column to your dataframe that displays what day of the week each date represents. You should search for how to do that using lubridate. Then write code to calculate the number of calls for each day of the week, and add a column to that result that calculates the percentage of all calls that occurred on each day of the week (so you want a dataframe with the day of the week, total number of calls and the percentage of calls on that day out of the total number of all calls). Describe your findings to me.

A2. Sunday had the most number of calls with a 15.1% of all calls. Following that was Saturday, Friday, Tuesday and Monday. All within the 14-15 percentage of the total. Weekends (Friday-Sunday) collectively make up roughly 45% of all overdose calls, suggesting a possible weekend effect where overdose occur more frequently toward the end of the week. Midweek days (Wed-Thursday) has the fewest calls, each below 14%

```{r}
baltcounty_calls <- baltcounty_calls |>
  mutate(day_of_week = wday(date, label = TRUE, abbr = FALSE))

calls_by_day <- baltcounty_calls |>
  group_by(day_of_week) |>
  summarise(num_calls = n()) |>
  mutate(percent_of_total = (num_calls/sum(num_calls)) * 100) |>
  arrange(desc(num_calls))

calls_by_day
```

Q3. Now let's look at locations. Which ones have the most calls? How would you describe them (feel free to search for more information on them)? Is there anything about the structure of the original data that might make you less confident in the counts by location or date?

A3. The data showed that 4540 Silver Spring Road has the most overdose calls, with 34 in total. Other top spots include police precincts like PC 02 on Windsor Mill Road and PC 06 on Susquehanna Avenue, as well as residential and commerical areas such as 330 Leanne Road and the Star Motel on Pulaski Highway. These results suggest that overdoses happen across a mix of residential neighborhoods, motels and possibly places where calls are just recorded, like police stations. That said, the numbers should be taken with a bit of caution. The way the data was entered isn't completely consistent --- some locations are full addresses, while others are intersections, business names or precinct offices. That means a few spots might look like major hotspots just because of how the data was logged. There could also be duplicates or slightly different versions of the same address. The data gives a general idea of where overdose calls are happening most often, but cleaning and standardizing the location information would help make the results more reliable. 

```{r}
calls_by_location <- baltcounty_calls |>
  group_by(location) |>
  summarise(num_calls = n()) |>
  arrange(desc(num_calls))

calls_by_location %>% slice_head(n = 10)
```

Q4. What's the best story idea or question you've seen as a result of the work you've done in this lab?

A4. I think one of the most interesting story ideas that came out of this lab is exploring why overdose calls seem to spike on the weekends and whether certain types of locations -- like motels, apartment complexes or specific neighborhoods -- see repeat incidents. It raises questions about what social or environmental factors contribute to these patterns. Are overdoses more common near major highways or areas with a dense population? Do weekend increases tie to social gatherings or access to certain substances? Looking deeper into these trends could lead to a strong data-driven story about how time and place could influence overdose risk, and what local health or law enforcement angencies are doing to respond to those patterns. 
