---
title: "lab_01"
author: "Aamaly Hossain"
date: "2025-09-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**). When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

Remember to add any new verbs or concepts to your class reference notebook!

## Load libraries and establish settings

**Task** Run the codeblock below to load the Tidyverse. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}
# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
```

## Load Data

We're going to load the UMD courses data.

**Task** Run the codeblock below to load the data, which is contained in an RDS file.

```{r}
umd_courses <- read_rds("lab_01.rds")
```

## Answer questions

**Q1** Which department has the highest mean (average) number of seats for its classes? You need to show the department with the highest average first. Which one is the most surprising/interesting to you? **A1** The Computer, Math, and Natural Sciences has the highest average number of seats for its classes -- with a whopping 440 seats. To be fair, there is only one class, but it still turns out to be the mean. 

```{r}
umd_courses |>
  group_by(department) |>
  summarise(
    count_classes = n(),
    total_seats = sum(seats),
    mean_seats = mean(seats),
  ) |>
  arrange(desc(mean_seats))

```

**Q2** Find out the greatest number of times any individual class has been offered in our dataset. Your code must show the greatest number of classes first.
**A2** Based on the code, it shows that CMSC798 has been offered 146 times, making it the greatest number of times any individual class has been offered.   

```{r}
umd_courses |>
  group_by(id) |>
  summarise(
    count_classes = n(),
    total_sections = sum(sections),
    maximum_sections = max(sections),
  ) |>
  arrange(desc(maximum_sections))
```

**Q3** Count the number of courses for each instructor and department. Which instructor has taught the most?
**A3** Keith Yearwood has taught 79 geographical science classes. There were no other instructors listed for the previous, and higher, numbers. 

```{r}
umd_courses |>
  group_by(instructors, department) |> 
  summarise(
    count_classes = n()
  ) |>
  arrange(desc(count_classes))
```

**Q4** Given the results for Q3, what would have made it easier to find the answer? You don't have to know *how* to accomplish this; I want you to describe in words what would make the answer easier to see. How confident are you that the answer you got is the right one, and why? **A4**

I'm not completely sure if the answer I got was the right one. But, I believe! I'm a bit doubtful of my answer choice because it seems like I could've found a higher number of courses that an instructor taught. I had to scroll to the seventh page, so it seemed like there is a better way to find the answer quicker. I think if I knew a code that would automatically get rid of all the "N/A" for the instructors, bascially cleaning all of that out -- then, I beleive I'd be pointed to the correct answer faster. 

**Q5** Take one of the questions you listed in the final question of the pre_lab and try to answer it here in code. If you don't know how to fully answer it, write what code you can and describe in words what else you would need to do. If you got an answer, is it interesting? **A5** I think I'd still have to figure out how to find the courses that have zero seats available (minimum), and how I can get that to be easily arranged, so it can be seen at the top row. Additionally, from what I've found, which I'm not too sure if it is correct, but University Courses have the most seats available within its section -- 2,876 seats. That figure is also the minimum.  

```{r}
#How many sections within each course have seats available? And what is that minimum and maximum number of seats available?  

umd_courses |>
  group_by(sections, department) |> 
  summarise(
    count_classes = n(),
    total_seats = sum(seats),
    minimum_seats = min(seats),
    maximum_seats = max(seats)
  ) |>
  arrange(desc(maximum_seats))

```
