---
title: "lab_09.Rmd"
author: "Aamaly Hossain"
date: "2025-11-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   A Datawrapper account
-   Data in CSV files in the `data` folder

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.

```{r}
# load libraries here
library(tidyverse)
library(janitor)
```

## Questions

**Q1.** You want to explore the differences between the electoral performances of Dan Cox in 2022 and Larry Hogan in 2018. Except a map is probably not the best way to show that. Let's make a chart instead, and the type is up to you, but the goal is to clearly show the differences between the two.

You MUST do the following things when answering this question: \* Load the md_gov_county.csv and md_gov_county_18.csv files from the data folder \* Join those dataframes, specifying the join column. \* Add columns to generate percentages for Cox and Hogan. \* Export that joined dataframe to a CSV file using `write_csv`. \* Using that exported CSV, make a chart with Datawrapper. What kind of chart is up to you, so experiment with a few kinds of charts to see what answers the question above in the clearest way. Your chart MUST: \* Have a title, description and source that tells readers about the data. The title and description should highlight the most newsworthy or important aspect. \* Easily readable labels and values \* Publish your chart \* Describe your thought process below - why you chose to make the chart you did - and add the URL of your chart.

**A1.** This may not be the most conventional, but I chose to make my chart this way because it was very easy to understand -- in my opinion. I found that in order to compare both Larry Hogan in 2018 and Dan Cox in 2022, a grouped bar chart was the way to go. It allowed for me to see the exact percent comparison for each person in each county. I also wanted to make sure that the counties which Hogan received the greatest percentage of votes would be in descending order because he had more votes than Cox in every county. So, the counties are not in alphabetical order, but in the order that gave Hogan the most support -- which tended to be Western Maryland and Eastern Shore. Additionally, I wanted to make sure the bars were different colors to show that Hogan won the 2018 gubernatorial election -- so his bar is red. And Dan Cox lost the 2022 gubernatorial election, but was apart of the Republic candidates in that race, so his bar is light yellow/orange. Even though it's a big long, it really shows the drastic difference between Hogan's votes in 2018 and Cox's votes in 2022. Chart URL:<https://datawrapper.dwcdn.net/2mCQO/2/>

```{r}
md_gov_county <- read_csv("data/md_gov_county.csv") |>
  clean_names()

colnames(md_gov_county)

md_gov_county_18 <- read_csv("data/md_gov_county_18.csv") |>
  clean_names()

colnames(md_gov_county_18)

joined_md_gov_county <- md_gov_county |> 
  left_join(md_gov_county_18, join_by(county))

joined_md_gov_county <- joined_md_gov_county |>
  mutate(
    total_votes = cox + moore + lashar + wallace + harding + sefcik + write_ins.x + write_ins.y + hogan + jealous + quinn + schlakman, 
    cox_pct = (cox / total_votes) * 100,
    hogan_pct = (hogan / total_votes) * 100
  )

write_csv(joined_md_gov_county, "data/joined_md_gov_county.csv")
```

**Q2.** You're interested in seeing how the methods of voting have changed in Maryland since 2018. Using the `md_votes_by_type.csv` that has data from the top of the ticket (governor in 2018 and 2022, president in 2020), make a chart showing the percentage of votes cast in each category - early voting, election day and mail - over the three years. You'll need to make those percentages in R and then export the CSV using `write_csv`, then make the chart in Datawrapper. Have a title, description and source that tells readers about the data. The title and description should highlight the most newsworthy or important aspect.

HINT: In Datawrapper's "Check and Describe" tab, you can click on a column to get the option of "hiding" it in any visualization.

What do you think the story here is? Put the answer below along with the URL of your chart.

**A2.** I think the story here is the significant decline of election day votes during the 2020 election. Whereas the 2018 and 2022 gubernatorial elections had significant election day votes, with limited mail-in ballots. In fact, the 2018 gubernatorial election had the least percentage of mail-in votes out of all versions of voting in the three difference races. While the 2018 and 2022 gubernatorial elections still saw most voters casting ballots in person on Election Day, 2020 stands out as an unprecedented shift toward mail voting, reflecting the impact of the pandemic and emergency election procedures. Another important point is how early voting has fluctuated across all three elections -- the least being in 2022 gubernatorial election. Chart URL: <https://datawrapper.dwcdn.net/Kp4mt/2/>

```{r}
md_votes_by_type <- read_csv("data/md_votes_by_type.csv") |>
  clean_names()

md_votes_by_type_pct <- md_votes_by_type |>
  mutate(
    early_pct = early / total * 100,
    election_day_pct = election_day / total * 100,
    mail_pct = mail / total * 100
  ) |>
  select(year, early_pct, election_day_pct, mail_pct)

write_csv(md_votes_by_type_pct, "data/md_votes_by_type_pct.csv")
```

**Q3.** You want to know more about the geography of 911 overdose calls in Montgomery County, Maryland, in 2022. To do that, you'll make a choropleth map in Datawrapper. You will need to:

-   Read the file into R and create columns for the number of calls in each zip code and the percentage of all calls that number represents and then export that CSV to use in your map.
-   Search for Maryland zip codes (use the 2020 zip codes) as the base map.
-   Upload the CSV file.
-   Change the tooltip to show both the number and the percentage (you can put them side by side or put one under the other). Make sure to properly format both numbers to make them readable.
-   Add labels to your map that locate prominent places in the county.
-   Have a title, description and source that tells readers about the data. The title and description should highlight the most newsworthy or important aspect.
-   Make any other changes to data that helps you make a more accurate map.

What do you think the story here is, and what would you want to know more about? Put the answer below along with the URL of your map.

**A3.** I think the story here would be the area that has the highest 911 overdose calls -- 20874. Germantown, Darnestown and Travilah are a few cities within that zipcode, so I think finding a commonality or neighborhoods within that zipcode would be beneficial. I'd want to know more about the specific neighborhoods where the 911 calls came from. I'd also like to know the demographics of the area with the highest amount of overdose calls and the lowest amount of overdose calls. Map URL:<https://datawrapper.dwcdn.net/gIBGv/1/>

```{r}
montgomery_2022_overdoses <- read.csv("data/montgomery_2022_overdoses.csv") |>
  clean_names()

overdose_by_zip <- montgomery_2022_overdoses |>
  filter(!is.na(zip)) |> 
  group_by(zip) |>
  summarise(
    calls = n()
  ) |>
  ungroup() |>
  mutate(
    total_calls = sum(calls),
    pct_calls = calls / total_calls * 100
  )

write_csv(overdose_by_zip, "data/overdoses_by_zip.csv")
```
